---
title: "palate_scRNA_zhao_2024_plot"
output: html_document
date: "2026-04-14"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  eval = FALSE,
  tidy.opts = list(width.cutoff = 120),
  message = TRUE,
  warning = FALSE,
  cache = TRUE,
  label.duplicate = "append"
)

```
This Rmd plot for Fig1a,b, and Supplementary Fig1 a-e

# 1 Load packages and datasets

```{r,load of packages and setup of the environment 1}
gc()
rm(list=ls())
library(Seurat)
library(data.table)
library(scales)
library(plotly)
library(ggplot2)
library(slingshot)
library(magrittr)
library(dplyr)
library(tidyverse)
library(cowplot)
library(gridExtra)
library(viridis)
library(readr)
library(Hmisc)
library(ggpubr)
library(rsample)
library(tidymodels)
library(scales)
library(tradeSeq)
library(pheatmap)
library(scCustomize)
#library(destiny)
library(plot3D)
library(rgl)
library(patchwork)
library(RColorBrewer)


set.seed(12)
theme_set(theme_bw())

Palate_zhao <- readRDS("./SeuratObject/20260511_palate_Zhao.rds")


```




## 1.1 Inspect datasets
### 1.1.1 Check cell percentage and stages


```{r, check cell percentage}
DefaultAssay(Palate_zhao) <- "ATAC"
table(Palate_zhao@meta.data[["celltype"]])
table(Palate_zhao@meta.data[["orig.ident"]])
```
```{r, check cell percentage}
DefaultAssay(Palate_zhao) <- "RNA"
table(Palate_zhao@meta.data[["celltype"]])
table(Palate_zhao@meta.data[["orig.ident"]])
```

## 1.1.2 Combine Developmental Stage
Now we add stage and unified parameters to this dataset
```{r}
DefaultAssay(Palate_zhao) <- "RNA"

Palate_zhao$Stage <- case_when(
  Palate_zhao$orig.ident %in% c("Zhao_E125-1", "Zhao_E125-2") ~ "E12.5",
  Palate_zhao$orig.ident %in% c("Zhao_E135-1", "Zhao_E135-2", "Zhao_E135-3") ~ "E13.5",
  Palate_zhao$orig.ident %in% c("Zhao_E140-1", "Zhao_E140-2") ~ "E14.0",
  Palate_zhao$orig.ident %in% c("Zhao_E145-1", "Zhao_E145-2") ~ "E14.5",
  TRUE ~ NA_character_
)

Palate_zhao$Dataset <- "Zhao"
#Palate_zhao$Anatomy <- "palate"
Palate_zhao$SampleID <- Palate_zhao$orig.ident
```

Inspect the staging
```{r}
table(Palate_zhao$orig.ident, Palate_zhao$Stage)
table(Palate_zhao$Stage, useNA = "ifany")
```

## 1.1.3 Check celltype distribution in different Stage
```{r, celltype count}
# 统计每个 Stage 中各 celltype 的细胞数
celltype_stage_count <- Palate_zhao@meta.data %>%
  group_by(Stage, celltype) %>%
  summarise(n_cells = n(), .groups = "drop") %>%
  arrange(Stage, desc(n_cells))

celltype_stage_count
```

```{r, check cell number in different stages}
p1 <- ggplot(celltype_stage_count, aes(x = Stage, y = n_cells, fill = celltype)) +
  geom_bar(stat = "identity", width = 0.75) +
  theme_classic(base_size = 14) +
  labs(
    x = "Stage",
    y = "Number of cells",
    fill = "Cell type"
  ) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "right"
  )

p1
```

```{r, export cell number count table}
# Convert to cell number dataframe
celltype_stage_matrix <- table(Palate_zhao$celltype, Palate_zhao$Stage)

celltype_stage_matrix

celltype_stage_matrix_df <- as.data.frame.matrix(celltype_stage_matrix)

celltype_stage_matrix_df

dir.create("./OutputTable", showWarnings = FALSE)

write.csv(
  celltype_stage_count,
  "./OutputTable/Palate_zhao_celltype_count_by_Stage_long.csv",
  row.names = FALSE
)

write.csv(
  celltype_stage_matrix_df,
  "./OutputTable/Palate_zhao_celltype_count_by_Stage_matrix.csv"
)
```




# 2 Basic plot
## 2.1 plot for scRNA
### 2.1.1 UMAP plot (Fig1a and Supp Fig1a)
```{r}
# 设置 RNA assay
DefaultAssay(Palate_zhao) <- "RNA"

# 设置细胞类型为身份
Idents(Palate_zhao) <- Palate_zhao$celltype

# 创建输出文件夹
dir.create("./OutputFigure", showWarnings = FALSE)

# 确认可用 reductions
Reductions(Palate_zhao)

# 使用 RNA Harmony UMAP
umap_use <- "umap.harmony"

# group_by_celltype
plot1 <- DimPlot(
  Palate_zhao,
  reduction = umap_use,
  pt.size = 0.01,
  label = TRUE,
  repel = TRUE
)

# group_by_orig.ident
plot2 <- DimPlot(
  Palate_zhao,
  reduction = umap_use,
  pt.size = 0.01,
  group.by = "orig.ident"
)

# group_by_Stage
plot3 <- DimPlot(
  Palate_zhao,
  reduction = umap_use,
  pt.size = 0.01,
  group.by = "Stage"
)

# split_by_orig.ident
plot4 <- DimPlot(
  Palate_zhao,
  reduction = umap_use,
  pt.size = 0.01,
  label = TRUE,
  repel = TRUE,
  split.by = "orig.ident"
)

# split_by_Stage
plot5 <- DimPlot(
  Palate_zhao,
  reduction = umap_use,
  pt.size = 0.01,
  label = TRUE,
  repel = TRUE,
  split.by = "Stage"
)

# combine: celltype + orig.ident + Stage
plotc <- plot1 + plot2 + plot3

ggsave(
  "./OutputFigure/Palate_zhao_umap_harmony_celltype_origident_Stage.pdf",
  plot = plotc,
  width = 30,
  height = 10
)

ggsave(
  "./OutputFigure/Palate_zhao_umap_harmony_split_by_origident.pdf",
  plot = plot4,
  width = 24,
  height = 10
)

ggsave(
  "./OutputFigure/Palate_zhao_umap_harmony_split_by_Stage.pdf",
  plot = plot5,
  width = 16,
  height = 10
)

# 单独查看
DimPlot(
  Palate_zhao,
  reduction = umap_use,
  pt.size = 0.01,
  label = TRUE,
  repel = TRUE
)

DimPlot(
  Palate_zhao,
  reduction = umap_use,
  pt.size = 0.01,
  group.by = "orig.ident"
)

DimPlot(
  Palate_zhao,
  reduction = umap_use,
  pt.size = 0.01,
  group.by = "Stage"
)
```

```{r}
library(Seurat)
library(ggplot2)

## 1. 设定 identity
Idents(Palate_zhao) <- "celltype"
# 使用 RNA Harmony UMAP
umap_use <- "umap.harmony"

## 2. 提取 Epithelial 细胞
cells_epi <- WhichCells(Palate_zhao, idents = "Epithelial")

## 3. 绘图
p_umap_epi_highlight <- DimPlot(
  Palate_zhao,
  reduction = umap_use,
  cells.highlight = cells_epi,
  cols = "grey85",
  cols.highlight = "#f8766d",
  pt.size = 0.15,
  sizes.highlight = 0.15
) +
  theme_classic() +
  theme(
    axis.line = element_line(color = "black"),
    axis.text = element_text(color = "black"),
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Whole palate scRNA-seq atlas",
    x = "UMAP_1",
    y = "UMAP_2"
  )

p_umap_epi_highlight

library(dplyr)

## 提取 UMAP 坐标
umap_df <- as.data.frame(Embeddings(Palate_zhao, umap_use))
umap_df$celltype <- Palate_zhao$celltype

## 计算每个 celltype 的标签位置
label_df <- umap_df %>%
  group_by(celltype) %>%
  summarise(
    UMAP_1 = median(UMAP_1),
    UMAP_2 = median(UMAP_2)
  )

## 重新作图并加标签
p_umap_epi_highlight_label <- DimPlot(
  Palate_zhao,
  reduction = umap_use,
  cells.highlight = cells_epi,
  cols = "grey85",
  cols.highlight = "#f8766d",
  pt.size = 0.15,
  sizes.highlight = 0.15
) +
  geom_text(
    data = label_df,
    aes(x = UMAP_1, y = UMAP_2, label = celltype),
    inherit.aes = FALSE,
    size = 6
  ) +
  theme_classic() +
  theme(
    axis.line = element_line(color = "black"),
    axis.text = element_text(color = "black"),
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Major cell classes in GSE218576 palate scRNA-seq",
    x = "UMAP_1",
    y = "UMAP_2"
  )

p_umap_epi_highlight_label

ggsave(
  "./OutputFigure/Palate_zhao_celltype_UMAP_Epithelial_highlight.pdf",
  plot = p_umap_epi_highlight_label,
  width = 6,
  height = 5.5
)
```

##2.2 Marker genes Dotplot  (Supp Fig1b)
```{r}
## Use celltype for marker identification
Idents(Palate_zhao) <- "celltype"

## Check celltype numbers
table(Idents(Palate_zhao))

## Perform DE using RNA rather than SCT
DefaultAssay(Palate_zhao) <- "RNA"

## Normalize RNA data if necessary
Palate_zhao <- NormalizeData(
  Palate_zhao,
  assay = "RNA",
  verbose = FALSE
)

## Scale RNA data
all.genes <- rownames(Palate_zhao)

Palate_zhao <- ScaleData(
  Palate_zhao,
  assay = "RNA",
  features = all.genes,
  verbose = FALSE
)

## Identify positive marker genes for each celltype
Palate_zhao_celltype_markers <- FindAllMarkers(
  object = Palate_zhao,
  assay = "RNA",
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

## Top 10 markers per celltype
top10_celltype <- Palate_zhao_celltype_markers %>%
  group_by(cluster) %>%
  slice_max(
    order_by = avg_log2FC,
    n = 10,
    with_ties = FALSE
  ) %>%
  ungroup()

## Top 6 markers using scCustomize
top6_celltype_markers <- Extract_Top_Markers(
  marker_dataframe = Palate_zhao_celltype_markers,
  num_genes = 6,
  named_vector = FALSE,
  make_unique = TRUE
)
```
```{r}
write.csv(
  Palate_zhao_celltype_markers,
  file = "./OutputTable/Palate_zhao_celltype_markers_RNA.csv",
  row.names = FALSE
)

write.csv(
  top10_celltype,
  file = "./OutputTable/Palate_zhao_celltype_top10_markers_RNA.csv",
  row.names = FALSE
)

write.csv(
  top6_celltype_markers,
  file = "./OutputTable/Palate_zhao_celltype_top6_markers_RNA.csv",
  row.names = FALSE
)
```




```{r}
## 1. 设置 marker genes
## 按照你附图中的 celltype marker 顺序
genes_to_check <- c(
  "Epcam", "Grhl2", "Cdh1",      # Epithelial
  "Prrx1", "Col1a1", "Twist1",     # Mesenchymal
  "Acta2", "Parm1", "Neb",       # Myogenic
  "Plxna4", "Grik2", "Sorcs1",        # Glial
  "Elmo1", "Ptprc",                # Myeloid
  "Cdh5", "Cldn5" ,  "Flt1",      # Endothelial
  "Hbb-y", "Hba-a1", "Hba-a2"      # Erythroid
)

## 去重，避免 DotPlot 重复 gene 报错或显示混乱
genes_to_check <- unique(genes_to_check)

## 检查基因是否存在
DefaultAssay(Palate_zhao) <- "RNA"

genes_present <- genes_to_check[genes_to_check %in% rownames(Palate_zhao)]
genes_missing <- genes_to_check[!genes_to_check %in% rownames(Palate_zhao)]

genes_present
genes_missing
```


```{r}
## =========================================================
## DotPlot for Palate_zhao whole object
## =========================================================


desired_order <- c(
  "Erythroid",
  "Endothelial",
  "Glial",
  "Myogenic",
  "Myeloid",
  "Mesenchymal",
  "Epithelial"
)

Palate_zhao$celltype <- factor(
  Palate_zhao$celltype,
  levels = desired_order
)

Idents(Palate_zhao) <- "celltype"

```

Extract RNA expression
```{r}
## 如果 RNA data layer 还没有 normalized data，先 NormalizeData
Palate_zhao <- NormalizeData(
  Palate_zhao,
  assay = "RNA",
  verbose = FALSE
)

exp_mat <- GetAssayData(
  object = Palate_zhao,
  assay = "RNA",
  layer = "data"
)

exp_mat <- exp_mat[genes_present, , drop = FALSE]

meta_use <- Palate_zhao@meta.data
meta_use$cell <- rownames(meta_use)

## 对齐 cell
common_cells <- intersect(colnames(exp_mat), rownames(meta_use))

exp_mat <- exp_mat[, common_cells, drop = FALSE]
meta_use <- meta_use[common_cells, , drop = FALSE]

meta_use$celltype <- factor(
  meta_use$celltype,
  levels = desired_order
)
```

Calculate average expression and percent expressed
```{r}
dot_df <- lapply(genes_present, function(gene_i) {
  
  exp_i <- as.numeric(exp_mat[gene_i, ])
  
  data.frame(
    cell = common_cells,
    gene = gene_i,
    expression = exp_i,
    celltype = meta_use$celltype
  )
}) %>%
  bind_rows() %>%
  filter(!is.na(celltype)) %>%
  group_by(celltype, gene) %>%
  summarise(
    avg_exp = mean(expression),
    pct_exp = mean(expression > 0) * 100,
    .groups = "drop"
  )

## 按 gene 对 average expression 做 scaling，模拟 Seurat DotPlot 的颜色逻辑
dot_df <- dot_df %>%
  group_by(gene) %>%
  mutate(
    avg_exp_scaled = as.numeric(scale(avg_exp))
  ) %>%
  ungroup()

## 限制极端值，类似 Seurat DotPlot 默认 scale.min / scale.max
dot_df$avg_exp_scaled[dot_df$avg_exp_scaled > 2.5] <- 2.5
dot_df$avg_exp_scaled[dot_df$avg_exp_scaled < -2.5] <- -2.5

dot_df$gene <- factor(
  dot_df$gene,
  levels = genes_present
)

dot_df$celltype <- factor(
  dot_df$celltype,
  levels = desired_order
)
```

Dotplot
```{r}
p_dot_Palate_zhao_marker_custom <- ggplot(
  dot_df,
  aes(
    x = gene,
    y = celltype
  )
) +
  geom_point(
    aes(
      size = pct_exp,
      color = avg_exp_scaled
    )
  ) +
  scale_size(
    range = c(0, 6),
    limits = c(0, 100),
    breaks = c(0, 25, 50, 75, 100),
    name = "Percent\nExpressed"
  ) +
  scale_color_gradientn(
    values = seq(0, 1, 0.2),
    colours = c("#295B9F", "#FFFFFF", "#D23B3E"),
    name = "Average\nExpression"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(
      angle = 45,
      hjust = 0.5,
      vjust = 0.5
    ),
    axis.text.y = element_text(size = 11),
    panel.border = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_blank()
  ) +
  labs(x = NULL, y = NULL) +
  guides(
    size = guide_legend(order = 3),
    color = guide_colorbar(order = 1)
  )

p_dot_Palate_zhao_marker_custom
```
```{r}
dir.create("./OutputFigure", showWarnings = FALSE, recursive = TRUE)

ggsave(
  "./OutputFigure/Palate_zhao_celltype_marker_dotplot_RNA.pdf",
  plot = p_dot_Palate_zhao_marker_custom,
  width = 9,
  height = 5
)
```


#3 Subset Epithelial cells
```{r}
DefaultAssay(Palate_zhao) <- "RNA"

Palate_zhao$Stage <- case_when(
  Palate_zhao$orig.ident %in% c("Zhao_E125-1", "Zhao_E125-2") ~ "E12.5",
  Palate_zhao$orig.ident %in% c("Zhao_E135-1", "Zhao_E135-2", "Zhao_E135-3") ~ "E13.5",
  Palate_zhao$orig.ident %in% c("Zhao_E140-1", "Zhao_E140-2") ~ "E14.0",
  Palate_zhao$orig.ident %in% c("Zhao_E145-1", "Zhao_E145-2") ~ "E14.5",
  TRUE ~ NA_character_
)

Palate_zhao$Dataset <- "Zhao"
Palate_zhao$SampleID <- Palate_zhao$orig.ident

Palate_zhao_epi <- subset(
  Palate_zhao,
  subset = celltype == "Epithelial"
)

Palate_zhao_epi$Anatomy <- "palatal_epithelium"
```


# 4 Remove ATAC and other unwanted metadata
```{r}
DefaultAssay(Palate_zhao_epi) <- "RNA"

if ("ATAC" %in% names(Palate_zhao_epi@assays)) {
  Palate_zhao_epi[["ATAC"]] <- NULL
}

Palate_zhao_epi@reductions <- list()
Palate_zhao_epi@graphs <- list()
Palate_zhao_epi@neighbors <- list()
```





# 5 SubCluster epithelial cells after harmony
## 5.1 SCT
```{r}
DefaultAssay(Palate_zhao_epi) <- "RNA"

# 如果还没有 percent.mt，先计算
if (!"percent.mt" %in% colnames(Palate_zhao_epi@meta.data)) {
  Palate_zhao_epi[["percent.mt"]] <- PercentageFeatureSet(
    Palate_zhao_epi,
    pattern = "^mt-"
  )
}

summary(Palate_zhao_epi$percent.mt)
sum(is.na(Palate_zhao_epi$percent.mt))

Palate_zhao_epi <- SCTransform(
  Palate_zhao_epi,
  assay = "RNA",
  new.assay.name = "SCT",
  vars.to.regress = "percent.mt",
  verbose = FALSE
)
```


```{r}
DefaultAssay(Palate_zhao_epi) <- "SCT"

Palate_zhao_epi <- RunPCA(
  Palate_zhao_epi,
  npcs = 50,
  features = VariableFeatures(Palate_zhao_epi)
)
```

## 5.2 Harmony
```{r}
system.time({
  Palate_zhao_epi <- RunHarmony(
    object = Palate_zhao_epi,
    assay.use = "SCT",
    reduction.use = "pca",
    dims.use = 1:50,
    group.by.vars = "orig.ident",
    plot_convergence = TRUE
  )
})
```

## 5.3 Check for stage
```{r}
table(Palate_zhao_epi$orig.ident, Palate_zhao_epi$Stage)
```

```{r, fig.height=5, fig.width= 20}
FeaturePlot(
  Palate_zhao_epi,
  features = "Krt14",
  split.by = "Stage"
)
```

```{r, fig.height=5, fig.width= 20}
FeaturePlot(
  Palate_zhao_epi,
  features = "Krt5",
  split.by = "Stage"
)
```

```{r, fig.height=5, fig.width= 20}
FeaturePlot(
  Palate_zhao_epi,
  features = "Krt6a",
  split.by = "Stage"
)
```

```{r,fig.width=15,fig.height=15}
FeaturePlot(Palate_zhao_epi, 
                  features = c("Krt14","Krt5","Krt6a","Krt4","Krt17"),
                  cols = c("grey85", brewer.pal(9,"YlOrRd")),
                  ncol=2, 
                  pt.size = 0.1,
                  min.cutoff = 0.1,
                  max.cutoff =5,
            label=T,repel = T)

```

## 5.4 Dimension reduction
```{r}
Palate_zhao_epi <- RunUMAP(
  object = Palate_zhao_epi,
  assay = "SCT",
  reduction = "harmony",
  dims = 1:50
)

Palate_zhao_epi <- FindNeighbors(
  object = Palate_zhao_epi,
  assay = "SCT",
  reduction = "harmony",
  dims = 1:50
)
```

```{r}
Palate_zhao_epi.resolution <- FindClusters(
  Palate_zhao_epi,
  dims = 1:50,
  graph.name = "SCT_snn",
  resolution = seq(from = 0, by = .2, length = 10)
)
```

```{r}
clustree(Palate_zhao_epi.resolution)
```

```{r}
DimPlot(
  Palate_zhao_epi.resolution,
  reduction = "umap",
  label = TRUE,
  group.by = "SCT_snn_res.0.2"
)

DimPlot(
  Palate_zhao_epi.resolution,
  reduction = "umap",
  label = TRUE,
  group.by = "SCT_snn_res.0.4"
)

DimPlot(
  Palate_zhao_epi.resolution,
  reduction = "umap",
  label = TRUE,
  group.by = "SCT_snn_res.0.6"
)

DimPlot(
  Palate_zhao_epi.resolution,
  reduction = "umap",
  label = TRUE,
  group.by = "SCT_snn_res.0.8"
)
```

Test Featureplot
```{r,fig.width=15,fig.height=15}
FeaturePlot(Palate_zhao_epi, 
                  features = c("Krt14","Krt5","Krt6a","Krt4","Krt17"),
                  cols = c("grey85", brewer.pal(9,"YlOrRd")),
                  ncol=2, 
                  pt.size = 0.1,
                  min.cutoff = 0.1,
                  max.cutoff =5,
            label=T,repel = T)

```
We choose 0.8 for reduction
```{r}
Palate_zhao_epi <- FindClusters(
  object = Palate_zhao_epi,
  graph.name = "SCT_snn",
  resolution = 0.8
)

```

## 5.5 UMAP plot for palate_epi (Fig1a and Supp Fig1c)
```{r}
# 设置 SCT assay
DefaultAssay(Palate_zhao_epi) <- "SCT"

# 设置细胞类型为身份
Idents(Palate_zhao_epi) <- Palate_zhao_epi$Celltype


# group_by_celltype
plot1 <- DimPlot(
  Palate_zhao_epi,
  reduction = "umap",
  pt.size = 0.01,
  label = TRUE,
  repel = TRUE
)

# group_by_orig.ident
plot2 <- DimPlot(
  Palate_zhao_epi,
  reduction = "umap",
  pt.size = 0.01,
  group.by = "orig.ident"
)

# group_by_Stage
plot3 <- DimPlot(
  Palate_zhao_epi,
  reduction = "umap",
  pt.size = 0.01,
  group.by = "Stage"
)

# split_by_orig.ident
plot4 <- DimPlot(
  Palate_zhao_epi,
  reduction = "umap",
  pt.size = 0.01,
  label = TRUE,
  repel = TRUE,
  split.by = "orig.ident"
)

# split_by_Stage
plot5 <- DimPlot(
  Palate_zhao_epi,
  reduction = "umap",
  pt.size = 0.01,
  label = TRUE,
  repel = TRUE,
  split.by = "Stage"
)

# combine: celltype + orig.ident + Stage
plotc <- plot1 + plot2 + plot3

ggsave(
  "./OutputFigure/Palate_zhao_epi_umap_harmony_celltype_origident_Stage.pdf",
  plot = plotc,
  width = 30,
  height = 10
)

ggsave(
  "./OutputFigure/Palate_zhao_epi_umap_harmony_split_by_origident.pdf",
  plot = plot4,
  width = 24,
  height = 10
)

ggsave(
  "./OutputFigure/Palate_zhao_epi_umap_harmony_split_by_Stage.pdf",
  plot = plot5,
  width = 16,
  height = 10
)

# 单独查看
DimPlot(
  Palate_zhao_epi,
  reduction = "umap",
  pt.size = 0.01,
  label = TRUE,
  repel = TRUE
)

DimPlot(
  Palate_zhao_epi,
  reduction = "umap",
  pt.size = 0.01,
  group.by = "orig.ident"
)

DimPlot(
  Palate_zhao_epi,
  reduction = "umap",
  pt.size = 0.01,
  group.by = "Stage"
)
```
# 6 PE celltype association and marker genes identification
## 6.1 Celltype association
merge clusters 
```{r}
DefaultAssay(Palate_zhao_epi) <- "SCT"
# remove previous celltype metadata
Palate_zhao_epi$celltype <- NULL
# Use resolution 0.8 as identity
Idents(Palate_zhao_epi) <- "SCT_snn_res.0.8"

# Check current clusters
table(Idents(Palate_zhao_epi))

# Remove cluster 16 before celltype assignment
Palate_zhao_epi <- subset(
  Palate_zhao_epi,
  idents = "16",
  invert = TRUE
)

# Check cluster 16 has been removed
table(Idents(Palate_zhao_epi))
```
```{r}
# Make sure identity is still SCT_snn_res.0.8
Idents(Palate_zhao_epi) <- "SCT_snn_res.0.8"

# Initialize Celltype
Palate_zhao_epi$Celltype <- NA_character_

# Assign Celltype based on SCT_snn_res.0.8
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "0"]  <- "PE0"
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "1"]  <- "PE1"
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "2"]  <- "PE2"
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "4"]  <- "PE3"
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "5"]  <- "PE4"
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "6"]  <- "PE5"
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "7"]  <- "PE6"
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "8"]  <- "PE7"
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "9"]  <- "PE8"
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "10"] <- "PE1"
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "11"] <- "PE9"
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "12"] <- "PE10"
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "13"] <- "PE11"
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "14"] <- "PE12"
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "17"] <- "PE9"
Palate_zhao_epi$Celltype[Palate_zhao_epi$SCT_snn_res.0.8 == "18"] <- "PE11"

# Set order
Palate_zhao_epi$Celltype <- factor(
  Palate_zhao_epi$Celltype,
  levels = c(
    "PE0", "PE1", "PE2", "PE3", "PE4", "PE5", "PE6",
    "PE7", "PE8", "PE9", "PE10", "PE11", "PE12"
  )
)

# Check assignment
table(Palate_zhao_epi$SCT_snn_res.0.8, Palate_zhao_epi$Celltype, useNA = "ifany")
table(Palate_zhao_epi$Celltype, useNA = "ifany")
```
```{r}
DimPlot(
  Palate_zhao_epi,
  reduction = "umap",
  group.by = "Celltype",
  label = TRUE,
  repel = TRUE,
  pt.size = 0.1
)
```

## 6.2 Marker genes identification
```{r}

# Use merged Celltype for marker identification
Idents(Palate_zhao_epi) <- "Celltype"

# Check celltype numbers
table(Idents(Palate_zhao_epi))
table(Palate_zhao_epi$Celltype, useNA = "ifany")

# Perform DE using RNA rather than SCT
DefaultAssay(Palate_zhao_epi) <- "RNA"

# Normalize RNA data if necessary
Palate_zhao_epi <- NormalizeData(Palate_zhao_epi)

# Scale RNA data
all.genes <- rownames(Palate_zhao_epi)

Palate_zhao_epi <- ScaleData(
  Palate_zhao_epi,
  features = all.genes
)

# Identify positive marker genes by Celltype
Palate_zhao_epi_Celltype_markers <- FindAllMarkers(
  object = Palate_zhao_epi,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

# Top 10 markers per Celltype
top10_Celltype <- Palate_zhao_epi_Celltype_markers %>%
  group_by(cluster) %>%
  slice_max(
    order_by = avg_log2FC,
    n = 10,
    with_ties = FALSE
  ) %>%
  ungroup()

# Top 6 markers using scCustomize
top6_Celltype_markers <- Extract_Top_Markers(
  marker_dataframe = Palate_zhao_epi_Celltype_markers,
  num_genes = 6,
  named_vector = FALSE,
  make_unique = TRUE
)
```

```{r}
write.csv(
  Palate_zhao_epi_Celltype_markers,
  file = "./OutputTable/Palate_zhao_epi_Celltype_markers.csv",
  row.names = FALSE
)

write.csv(
  top10_Celltype,
  file = "./OutputTable/Palate_zhao_epi_Celltype_top10_markers.csv",
  row.names = FALSE
)

write.csv(
  top6_Celltype_markers,
  file = "./OutputTable/Palate_zhao_epi_Celltype_top6_markers.csv",
  row.names = FALSE
)
```

```{r}
Idents(Palate_zhao_epi) <- "Celltype"
DefaultAssay(Palate_zhao_epi) <- "RNA"

p_heatmap_Celltype <- DoHeatmap(
  Palate_zhao_epi,
  features = unique(top10_Celltype$gene),
  group.by = "Celltype"
)

p_heatmap_Celltype

ggsave(
  "./OutputFigure/Palate_zhao_epi_Celltype_markers_top10_heatmap.pdf",
  plot = p_heatmap_Celltype,
  device = "pdf",
  width = 30,
  height = 40,
  units = "cm"
)
```

## 6.3 Celltype marker genes FeaturePlot
```{r}
top5_Celltype <- Palate_zhao_epi_Celltype_markers %>%
  group_by(cluster) %>%
  slice_max(
    order_by = avg_log2FC,
    n = 5,
    with_ties = FALSE
  ) %>%
  ungroup()

write.csv(
  top5_Celltype,
  file = "./OutputTable/Palate_zhao_epi_Celltype_top5_markers.csv",
  row.names = FALSE
)
```


```{r}
plot_top5_marker_with_highlight_Celltype <- function(
    obj,
    marker_df,
    assay_use = "RNA",
    celltype_col = "Celltype",
    output_prefix = "Palate_zhao_epi_Celltype"
) {
  
  DefaultAssay(obj) <- assay_use
  Idents(obj) <- celltype_col
  
  celltype_use <- unique(as.character(marker_df$cluster))
  celltype_use <- celltype_use[
    order(as.numeric(gsub("PE", "", celltype_use)))
  ]
  
  for (i_chr in celltype_use) {
    
    genes_to_plot <- marker_df$gene[as.character(marker_df$cluster) == i_chr]
    genes_to_plot <- unique(genes_to_plot)
    genes_to_plot <- genes_to_plot[genes_to_plot %in% rownames(obj[[assay_use]])]
    
    cells_i <- WhichCells(obj, idents = i_chr)
    
    if (length(genes_to_plot) > 0 && length(cells_i) > 0) {
      
      p_highlight <- DimPlot(
        obj,
        reduction = "umap",
        cells.highlight = list(cells_i),
        cols = "grey85",
        cols.highlight = "#D23B3E",
        pt.size = 0.1,
        sizes.highlight = 0.25
      ) +
        ggtitle(paste0("Highlight: ", i_chr)) +
        NoLegend() +
        theme(
          plot.title = element_text(hjust = 0.5, face = "bold")
        )
      
      p_features <- FeaturePlot(
        obj,
        features = genes_to_plot,
        reduction = "umap",
        cols = c("grey85", brewer.pal(9, "YlOrRd")),
        pt.size = 0.1,
        order = TRUE,
        combine = FALSE
      )
      
      p_features <- lapply(
        seq_along(p_features),
        function(j) {
          p_features[[j]] +
            ggtitle(genes_to_plot[j]) +
            theme(
              plot.title = element_text(hjust = 0.5, face = "bold")
            )
        }
      )
      
      p_all <- wrap_plots(
        c(list(p_highlight), p_features),
        ncol = 2
      ) +
        plot_annotation(
          title = paste0(
            "Top 5 marker genes in ",
            i_chr,
            " (",
            assay_use,
            ")"
          )
        )
      
      print(p_all)
      
      ggsave(
        filename = paste0(
          "./OutputFigure/",
          output_prefix,
          "_",
          i_chr,
          "_top5_marker_plus_highlight_",
          assay_use,
          ".pdf"
        ),
        plot = p_all,
        width = 10,
        height = 12
      )
      
    } else {
      print(paste("No genes or no cells to plot for", i_chr))
    }
  }
}
```

```{r}
plot_top5_marker_with_highlight_Celltype(
  obj = Palate_zhao_epi,
  marker_df = top5_Celltype,
  assay_use = "RNA",
  celltype_col = "Celltype",
  output_prefix = "Palate_zhao_epi_Celltype"
)
```

```{r}
plot_top5_marker_with_highlight_Celltype(
  obj = Palate_zhao_epi,
  marker_df = top5_Celltype,
  assay_use = "SCT",
  celltype_col = "Celltype",
  output_prefix = "Palate_zhao_epi_Celltype"
)
```

## 6.4 Celltype marker genes Dotplot (Supp Fig1d)



Set genes of interest
```{r}
## 1. Set marker genes
genes_to_check <- c(
  "Cacna1e", "Fstl4",       # PE0
  "Nell2", "Unc13c",        # PE1
  "Krt6a", "Zfp750",        # PE2
  "Pif1", "Ube2c",          # PE3
  "Maf", "Dab1",            # PE4
  "H2ac10", "H2bc18",       # PE5
  "Abi3bp", "Ccdc3",        # PE6
  "Bmper", "Sostdc1",       # PE7
  "Olfm5", "Tspan8",        # PE8
  "Rp1", "Zbbx",            # PE9
  "Ccnb2", "Knstrn",        # PE10
  "Prdm16", "Rarb",         # PE11
  "Nrarp", "Palm3"          # PE12
)

genes_to_check <- unique(genes_to_check)

## 2. Check if genes exist
DefaultAssay(Palate_zhao_epi) <- "SCT"

genes_present <- genes_to_check[genes_to_check %in% rownames(Palate_zhao_epi)]
genes_missing <- genes_to_check[!genes_to_check %in% rownames(Palate_zhao_epi)]

genes_present
genes_missing
```
Set Celltype order
```{r}
## =========================================================
## Set Celltype order
## =========================================================

desired_order <- rev(paste0("PE", 0:12))

Palate_zhao_epi$Celltype <- factor(
  Palate_zhao_epi$Celltype,
  levels = desired_order
)

Idents(Palate_zhao_epi) <- "Celltype"

table(Idents(Palate_zhao_epi))
```
Extract SCT expression matrix
```{r}
## =========================================================
## Extract SCT expression matrix
## =========================================================

exp_mat <- GetAssayData(
  object = Palate_zhao_epi,
  assay = "SCT",
  layer = "data"
)

exp_mat <- exp_mat[genes_present, , drop = FALSE]

meta_use <- Palate_zhao_epi@meta.data
meta_use$cell <- rownames(meta_use)

## 对齐 cell
common_cells <- intersect(colnames(exp_mat), rownames(meta_use))

exp_mat <- exp_mat[, common_cells, drop = FALSE]
meta_use <- meta_use[common_cells, , drop = FALSE]

meta_use$Celltype <- factor(
  meta_use$Celltype,
  levels = desired_order
)
```

Calculate average expression and percent expressed
```{r}
## =========================================================
## Calculate average expression and percent expressed
## =========================================================

dot_df <- lapply(genes_present, function(gene_i) {
  
  exp_i <- as.numeric(exp_mat[gene_i, ])
  
  data.frame(
    cell = common_cells,
    gene = gene_i,
    expression = exp_i,
    Celltype = meta_use$Celltype
  )
}) %>%
  bind_rows() %>%
  filter(!is.na(Celltype)) %>%
  group_by(Celltype, gene) %>%
  summarise(
    avg_exp = mean(expression),
    pct_exp = mean(expression > 0) * 100,
    .groups = "drop"
  )

## 按 gene 对 average expression 做 scaling，模拟 Seurat DotPlot 颜色逻辑
dot_df <- dot_df %>%
  group_by(gene) %>%
  mutate(
    avg_exp_scaled = as.numeric(scale(avg_exp))
  ) %>%
  ungroup()

## 如果某个 gene 在所有 Celltype 平均值完全一样，scale 会产生 NA，改成 0
dot_df$avg_exp_scaled[is.na(dot_df$avg_exp_scaled)] <- 0

## 限制极端值
dot_df$avg_exp_scaled[dot_df$avg_exp_scaled > 2.5] <- 2.5
dot_df$avg_exp_scaled[dot_df$avg_exp_scaled < -2.5] <- -2.5

dot_df$gene <- factor(
  dot_df$gene,
  levels = genes_present
)

dot_df$Celltype <- factor(
  dot_df$Celltype,
  levels = desired_order
)
```

Dotplot
```{r}
## =========================================================
## DotPlot
## =========================================================

p_dot_Palate_zhao_epi_marker_custom <- ggplot(
  dot_df,
  aes(
    x = gene,
    y = Celltype
  )
) +
  geom_point(
    aes(
      size = pct_exp,
      color = avg_exp_scaled
    )
  ) +
  scale_size(
    range = c(0, 6),
    limits = c(0, 100),
    breaks = c(0, 25, 50, 75, 100),
    name = "Percent\nExpressed"
  ) +
  scale_color_gradientn(
    values = seq(0, 1, 0.2),
    colours = c("#295B9F", "#FFFFFF", "#D23B3E"),
    name = "Average\nExpression"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(
      angle = 45,
      hjust = 0.5,
      vjust = 0.5
    ),
    axis.text.y = element_text(size = 11),
    panel.border = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_blank()
  ) +
  labs(x = NULL, y = NULL) +
  guides(
    size = guide_legend(order = 3),
    color = guide_colorbar(order = 1)
  )

p_dot_Palate_zhao_epi_marker_custom
```
Save
```{r}
dir.create("./OutputFigure", showWarnings = FALSE, recursive = TRUE)

ggsave(
  "./OutputFigure/Palate_zhao_epi_Celltype_marker_dotplot_SCT_custom.pdf",
  plot = p_dot_Palate_zhao_epi_marker_custom,
  width = 8.5,
  height = 5
)

ggsave(
  "./OutputFigure/Palate_zhao_epi_Celltype_marker_dotplot_SCT_custom.png",
  plot = p_dot_Palate_zhao_epi_marker_custom,
  width = 8.5,
  height = 5,
  dpi = 300
)
```


Then I need to add the basal progenitor markers in embryonic skin scRNA

Set genes of interest
```{r}
## 1. Set marker genes
genes_to_check <- c(
  # PE0
  "Cacna1e", "Fstl4",
  
  # PE1
  "Nell2", "Unc13c",
  
  # PE2 / periderm
  "Krt6a", "Zfp750",
  
  # PE3
  "Pif1", "Ube2c",
  
  # PE4
  "Maf", "Dab1",
  
  # PE5
  "H2ac10", "H2bc18",
  
  # PE6
  "Abi3bp", "Ccdc3",
  
  # PE7
  "Bmper", "Sostdc1",
  
  # PE8
  "Olfm5", "Tspan8",
  
  # PE9
  "Rp1", "Zbbx",
  
  # PE10
  "Ccnb2", "Knstrn",
  
  # PE11
  "Prdm16", "Rarb",
  
  # PE12
  "Nrarp", "Palm3",
  
  # PE12
  "Krt5", "Krt14",
  
  # Jacob et al. BasalTagln-associated genes
  "C1qtnf3", "Hapln1", "Tagln", "Cldn11",
  "Amtn", "Myl9", "Bambi", "Wnt2", "Wnt3a", "Wnt4", "Wnt6"
)

genes_to_check <- unique(genes_to_check)

## 2. Check if genes exist
DefaultAssay(Palate_zhao_epi) <- "SCT"

genes_present <- genes_to_check[genes_to_check %in% rownames(Palate_zhao_epi)]
genes_missing <- genes_to_check[!genes_to_check %in% rownames(Palate_zhao_epi)]

genes_present
genes_missing
```
Set Celltype order
```{r}
## =========================================================
## Set Celltype order
## =========================================================

desired_order <- rev(paste0("PE", 0:12))

Palate_zhao_epi$Celltype <- factor(
  Palate_zhao_epi$Celltype,
  levels = desired_order
)

Idents(Palate_zhao_epi) <- "Celltype"

table(Idents(Palate_zhao_epi))
```
Extract SCT expression matrix
```{r}
## =========================================================
## Extract SCT expression matrix
## =========================================================

exp_mat <- GetAssayData(
  object = Palate_zhao_epi,
  assay = "SCT",
  layer = "data"
)

exp_mat <- exp_mat[genes_present, , drop = FALSE]

meta_use <- Palate_zhao_epi@meta.data
meta_use$cell <- rownames(meta_use)

## 对齐 cell
common_cells <- intersect(colnames(exp_mat), rownames(meta_use))

exp_mat <- exp_mat[, common_cells, drop = FALSE]
meta_use <- meta_use[common_cells, , drop = FALSE]

meta_use$Celltype <- factor(
  meta_use$Celltype,
  levels = desired_order
)
```

Calculate average expression and percent expressed
```{r}
## =========================================================
## Calculate average expression and percent expressed
## =========================================================

dot_df <- lapply(genes_present, function(gene_i) {
  
  exp_i <- as.numeric(exp_mat[gene_i, ])
  
  data.frame(
    cell = common_cells,
    gene = gene_i,
    expression = exp_i,
    Celltype = meta_use$Celltype
  )
}) %>%
  bind_rows() %>%
  filter(!is.na(Celltype)) %>%
  group_by(Celltype, gene) %>%
  summarise(
    avg_exp = mean(expression),
    pct_exp = mean(expression > 0) * 100,
    .groups = "drop"
  )

## 按 gene 对 average expression 做 scaling，模拟 Seurat DotPlot 颜色逻辑
dot_df <- dot_df %>%
  group_by(gene) %>%
  mutate(
    avg_exp_scaled = as.numeric(scale(avg_exp))
  ) %>%
  ungroup()

## 如果某个 gene 在所有 Celltype 平均值完全一样，scale 会产生 NA，改成 0
dot_df$avg_exp_scaled[is.na(dot_df$avg_exp_scaled)] <- 0

## 限制极端值
dot_df$avg_exp_scaled[dot_df$avg_exp_scaled > 2.5] <- 2.5
dot_df$avg_exp_scaled[dot_df$avg_exp_scaled < -2.5] <- -2.5

dot_df$gene <- factor(
  dot_df$gene,
  levels = genes_present
)

dot_df$Celltype <- factor(
  dot_df$Celltype,
  levels = desired_order
)
```

Dotplot
```{r}
## =========================================================
## DotPlot
## =========================================================

p_dot_Palate_zhao_epi_marker_custom <- ggplot(
  dot_df,
  aes(
    x = gene,
    y = Celltype
  )
) +
  geom_point(
    aes(
      size = pct_exp,
      color = avg_exp_scaled
    )
  ) +
  scale_size(
    range = c(0, 6),
    limits = c(0, 100),
    breaks = c(0, 25, 50, 75, 100),
    name = "Percent\nExpressed"
  ) +
  scale_color_gradientn(
    values = seq(0, 1, 0.2),
    colours = c("#295B9F", "#FFFFFF", "#D23B3E"),
    name = "Average\nExpression"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(
      angle = 45,
      hjust = 0.5,
      vjust = 0.5
    ),
    axis.text.y = element_text(size = 11),
    panel.border = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_blank()
  ) +
  labs(x = NULL, y = NULL) +
  guides(
    size = guide_legend(order = 3),
    color = guide_colorbar(order = 1)
  )

p_dot_Palate_zhao_epi_marker_custom
```
Save
```{r}
dir.create("./OutputFigure", showWarnings = FALSE, recursive = TRUE)

ggsave(
  "./OutputFigure/Palate_zhao_epi_Celltype_marker_w_embryonic_skin_progenitor_markers_dotplot_SCT_custom_with_k14K5.pdf",
  plot = p_dot_Palate_zhao_epi_marker_custom,
  width = 10,
  height = 5
)

ggsave(
  "./OutputFigure/Palate_zhao_epi_Celltype_marker_w_embryonic_skin_progenitor_markers_dotplot_SCT_custom_with_k14K5.png",
  plot = p_dot_Palate_zhao_epi_marker_custom,
  width = 10,
  height = 5,
  dpi = 300
)
```


# 7 PE cell percentage changes (Supp Fig1e)

```{r}
## 设置顺序
Palate_zhao_epi$Stage <- factor(
  Palate_zhao_epi$Stage,
  levels = c("E12.5", "E13.5", "E14.0", "E14.5")
)

Palate_zhao_epi$Celltype <- factor(
  Palate_zhao_epi$Celltype,
  levels = paste0("PE", 12:0)
)

## 计算每个 Stage 内不同 Celltype 的百分比
celltype_stage_percent <- Palate_zhao_epi@meta.data %>%
  filter(!is.na(Stage), !is.na(Celltype)) %>%
  count(Stage, Celltype, name = "cell_number") %>%
  complete(
    Stage,
    Celltype,
    fill = list(cell_number = 0)
  ) %>%
  group_by(Stage) %>%
  mutate(
    stage_total = sum(cell_number),
    percentage = cell_number / stage_total * 100
  ) %>%
  ungroup()

celltype_stage_percent

write.csv(
  celltype_stage_percent,
  file = "./OutputTable/Palate_zhao_epi_Celltype_percentage_by_Stage_barplot_rawdata.csv",
  row.names = FALSE
)
```


Barplot to show cell percentage changes through time

```{r}
p_celltype_percent_stage_dodge <- ggplot(
  celltype_stage_percent,
  aes(
    x = Celltype,
    y = percentage,
    fill = Stage
  )
) +
  geom_bar(
    stat = "identity",
    position = position_dodge(width = 0.8),
    width = 0.7,
    color = "black",
    linewidth = 0.2
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x = "Celltype",
    y = "Percentage within each stage",
    fill = "Stage"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

p_celltype_percent_stage_dodge

ggsave(
  "./OutputFigure/Palate_zhao_epi_Celltype_percentage_by_Stage_dodge_barplot.pdf",
  plot = p_celltype_percent_stage_dodge,
  width = 10,
  height = 3,
  units = "in"
)

ggsave(
  "./OutputFigure/Palate_zhao_epi_Celltype_percentage_by_Stage_dodge_barplot.png",
  plot = p_celltype_percent_stage_dodge,
  width = 10,
  height = 3,
  units = "in",
  dpi = 300
)
```

#8 Slingshot pseudotime trajectories
## 8.1 Generate slingshot object
```{r}
Palate_zhao_epi.sc <- as.SingleCellExperiment(Palate_zhao_epi)
table(Palate_zhao_epi.sc$ident)
reducedDim(Palate_zhao_epi.sc, "UMAP") <- Embeddings(Palate_zhao_epi, "umap")
sce <- slingshot(Palate_zhao_epi.sc, clusterLabels = Palate_zhao_epi.sc$ident, reducedDim = "UMAP", allow.breaks = FALSE)

sce
```

```{r}
sce <- slingshot(
  Palate_zhao_epi.sc,
  clusterLabels = Palate_zhao_epi.sc$ident,
  reducedDim = "UMAP",
  end.clus = "PE2",
  allow.breaks = FALSE
)
slingLineages(sce) 
```

$Lineage1 [1] "PE12" "PE2"
$Lineage2 [1] "PE4", "PE1", "PE6", "PE11", "PE8", "PE2"

From the above interpretation, there are two trajectories towards PE2 (periderm)

Now plot Fig 1b
## 8.2 Plot Fig1b

Define function for trajectory ploting
```{r}
plot_slingshot_lineage <- function(
    lineage_index,
    lineage_title,
    sce_object,
    plot_data,
    point_size = 0.35,
    background_size = 0.25,
    curve_width = 1.2,
    show_labels = TRUE
) {

  pt_col <- paste0("Lineage", lineage_index)

  lineage_clusters <- slingLineages(sce_object)[[lineage_index]]

  df <- plot_data

  # 将每条 lineage 的 pseudotime 单独缩放至 0–1
  df$pseudotime <- df[[pt_col]]

  use_cells <- !is.na(df$pseudotime)

  df$pseudotime_scaled <- NA_real_
  df$pseudotime_scaled[use_cells] <- scales::rescale(
    df$pseudotime[use_cells],
    to = c(0, 1)
  )

  # 当前 lineage 的 principal curve
  current_curve <- slingCurves(sce_object)[[lineage_index]]

  curve_xy <- current_curve$s

  # 根据 principal curve 的顺序整理坐标
  if (!is.null(current_curve$ord)) {
    curve_xy <- curve_xy[current_curve$ord, , drop = FALSE]
  }

  curve_df <- data.frame(
    UMAP_1 = curve_xy[, 1],
    UMAP_2 = curve_xy[, 2]
  )

  # 计算需要标注的 cluster 中心位置
  label_df <- do.call(
    rbind,
    lapply(lineage_clusters, function(cl) {

      cells_this_cluster <- df$cluster == cl & use_cells

      if (!any(cells_this_cluster)) {
        return(NULL)
      }

      data.frame(
        cluster = cl,
        UMAP_1 = median(df$UMAP_1[cells_this_cluster]),
        UMAP_2 = median(df$UMAP_2[cells_this_cluster])
      )
    })
  )

  p <- ggplot(df, aes(x = UMAP_1, y = UMAP_2)) +

    # 所有细胞作为灰色背景
    geom_point(
      color = "grey88",
      size = background_size
    ) +

    # 当前 lineage 细胞按 pseudotime 着色
    geom_point(
      data = df[use_cells, , drop = FALSE],
      aes(color = pseudotime_scaled),
      size = point_size
    ) +

    # Slingshot 曲线
    geom_path(
      data = curve_df,
      aes(x = UMAP_1, y = UMAP_2),
      inherit.aes = FALSE,
      color = "black",
      linewidth = curve_width,
      lineend = "round"
    ) +

    scale_color_viridis_c(
      option = "D",
      direction = 1,
      limits = c(0, 1),
      breaks = c(0, 0.25, 0.50, 0.75, 1),
      name = "Pseudotime"
    ) +

    coord_fixed() +

    labs(
      title = lineage_title,
      x = "UMAP_1",
      y = "UMAP_2"
    ) +

    theme_classic(base_size = 12) +

    theme(
      plot.title = element_text(
        hjust = 0.5,
        face = "plain",
        size = 13
      ),
      legend.title = element_text(size = 11),
      legend.text = element_text(size = 9),
      axis.title = element_text(size = 11)
    )

  if (show_labels) {
    p <- p +
      geom_text(
        data = label_df,
        aes(
          x = UMAP_1,
          y = UMAP_2,
          label = cluster
        ),
        inherit.aes = FALSE,
        color = "black",
        size = 4,
        fontface = "plain",
        nudge_y = 0.5
      )
  }

  return(p)
}

```
Extract UMAP pseudotime and curve

```{r}
umap_mat <- reducedDim(sce, "UMAP")

pt_mat <- slingPseudotime(sce, na = TRUE)
curve_list <- slingCurves(sce)

dim(pt_mat)
colnames(pt_mat)
slingLineages(sce)

```

```{r}
plot_df <- data.frame(
  cell = rownames(umap_mat),
  UMAP_1 = umap_mat[, 1],
  UMAP_2 = umap_mat[, 2],
  cluster = as.character(Palate_zhao_epi.sc$ident),
  Lineage1 = pt_mat[, 1],
  Lineage2 = pt_mat[, 2],
  stringsAsFactors = FALSE
)
```


```{r}
p_traj2 <- plot_slingshot_lineage(
  lineage_index = 1,
  lineage_title = "Periderm-related trajectory 2",
  sce_object = sce,
  plot_data = plot_df
)

p_traj1 <- plot_slingshot_lineage(
  lineage_index = 2,
  lineage_title = "Periderm-related trajectory 1",
  sce_object = sce,
  plot_data = plot_df
)

p_combined <- p_traj1 + p_traj2 +
  plot_layout(
    widths = c(1, 1),
    guides = "collect"
  ) &
  theme(
    legend.position = "right"
  )

p_combined

ggsave(
  filename = "./OutputFigure/Palate_zhao_epi_slingshot_two_lineages.pdf",
  plot = p_combined,
  width = 10,
  height = 4.5,
  useDingbats = FALSE
)
```

Modify the combined plot with Adobe Illustrator.

#9 Plot Krt genes (Fig1a)

```{r}
cutoff_Krt5  <- 0.5
cutoff_Krt14 <- 0.5
cutoff_Krt6a <- 1

data_Krt5 <- data_gene[, c("cell", "UMAP_1", "UMAP_2", "Krt5")]
data_Krt5 <- data_Krt5[data_Krt5$Krt5 > cutoff_Krt5, ]

data_Krt14 <- data_gene[, c("cell", "UMAP_1", "UMAP_2", "Krt14")]
data_Krt14 <- data_Krt14[data_Krt14$Krt14 > cutoff_Krt14, ]

data_Krt6a <- data_gene[, c("cell", "UMAP_1", "UMAP_2", "Krt6a")]
data_Krt6a <- data_Krt6a[data_Krt6a$Krt6a > cutoff_Krt6a, ]

point_size_bg <- 0.28
point_alpha_bg <- 0.35

point_size_Krt14 <- 0.95
point_alpha_Krt14 <- 0.75

point_size_Krt5 <- 0.90
point_alpha_Krt5 <- 1.00

point_size_Krt6a <- 0.95
point_alpha_Krt6a <- 1.00

library(ggplot2)
library(ggnewscale)
library(scales)

p_Krt5_Krt14_Krt6a <- ggplot() +
  ## 背景
  geom_point(
    data = data_gene,
    aes(x = UMAP_1, y = UMAP_2),
    size = point_size_bg,
    shape = 16,
    color = "grey90",
    alpha = point_alpha_bg
  ) +
  
  ## 先画 Krt14：恢复成明显红色
  new_scale_color() +
  geom_point(
    data = data_Krt14,
    aes(x = UMAP_1, y = UMAP_2, color = Krt14),
    shape = 16,
    size = point_size_Krt14,
    alpha = point_alpha_Krt14
  ) +
  scale_color_gradient(
    low = "grey85",
    high = "#006400",
    limits = c(cutoff_Krt14, 5),
    oob = scales::squish,
    name = "Krt14"
  ) +
  
  ## 再画 Krt5：深蓝、更明显
  new_scale_color() +
  geom_point(
    data = data_Krt5,
    aes(x = UMAP_1, y = UMAP_2, color = Krt5),
    shape = 16,
    size = point_size_Krt5,
    alpha = point_alpha_Krt5
  ) +
  scale_color_gradient(
    low = "grey85",
    high = "#B30000",
    limits = c(cutoff_Krt5, 5),
    oob = scales::squish,
    name = "Krt5"
  ) +
  
  ## 最后画 Krt6a：深绿、最明显
  new_scale_color() +
  geom_point(
    data = data_Krt6a,
    aes(x = UMAP_1, y = UMAP_2, color = Krt6a),
    shape = 16,
    size = point_size_Krt6a,
    alpha = point_alpha_Krt6a
  ) +
  scale_color_gradient(
    low = "grey85",
    high = "#003B8E",
    limits = c(cutoff_Krt6a, 5),
    oob = scales::squish,
    name = "Krt6a"
  ) +
  
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "right"
  ) +
  labs(
    x = "UMAP_1",
    y = "UMAP_2",
    title = "Krt5, Krt14 and Krt6a expression in Palate_zhao_epi"
  )
p_Krt5_Krt14_Krt6a <- p_Krt5_Krt14_Krt6a +
  coord_fixed()

p_Krt5_Krt14_Krt6a
```

```{r}
ggsave(
  "./OutputFigure/Palate_zhao_epi_SCT_Krt5_Krt14_Krt6a_overlay_UMAP_final.pdf",
  plot = p_Krt5_Krt14_Krt6a,
  width = 6,
  height = 5.5,
  units = "in"
)
```


# 0 Save Palate_zhao_epi object
```{r}
saveRDS(
  Palate_zhao_epi,
  file = "./SeuratObject/Palate_zhao_epi_Celltype_marker_analysis.rds"
)
```

# 1 Extract Krt14 and Krt5 expression matrix
## 9.1 Extract normalized RNA matrix
```{r}
genes_use <- c("Krt5", "Krt14")

## 检查 RNA 中是否存在
DefaultAssay(Palate_zhao_epi) <- "RNA"

genes_use[genes_use %in% rownames(Palate_zhao_epi[["RNA"]])]
genes_use[!genes_use %in% rownames(Palate_zhao_epi[["RNA"]])]
```

```{r}

Palate_zhao_epi <- NormalizeData(
  Palate_zhao_epi,
  assay = "RNA",
  verbose = FALSE
)

## 提取 RNA normalized expression
expr_RNA <- FetchData(
  Palate_zhao_epi,
  vars = genes_use,
  assay = "RNA",
  layer = "data"
)

expr_RNA$cellID <- rownames(expr_RNA)

## 加 metadata
meta_use <- Palate_zhao_epi@meta.data
meta_use$cellID <- rownames(meta_use)

cols_meta <- c("cellID", "Celltype", "Stage", "orig.ident")
cols_meta <- cols_meta[cols_meta %in% colnames(meta_use)]

expr_RNA_out <- expr_RNA %>%
  left_join(
    meta_use[, cols_meta, drop = FALSE],
    by = "cellID"
  ) %>%
  select(cellID, Krt5, Krt14, everything())

write.csv(
  expr_RNA_out,
  file = "./OutputTable/Palate_zhao_epi_Krt5_Krt14_RNA_expression.csv",
  row.names = FALSE,
  quote = FALSE
)

head(expr_RNA_out)
```

## 9.2 Extract SCT data
```{r}
DefaultAssay(Palate_zhao_epi) <- "SCT"

## 检查 SCT 中是否存在
genes_use[genes_use %in% rownames(Palate_zhao_epi[["SCT"]])]
genes_use[!genes_use %in% rownames(Palate_zhao_epi[["SCT"]])]
```

```{r}
## 提取 SCT data layer expression
expr_SCT <- FetchData(
  Palate_zhao_epi,
  vars = genes_use,
  assay = "SCT",
  layer = "data"
)

expr_SCT$cellID <- rownames(expr_SCT)

expr_SCT_out <- expr_SCT %>%
  left_join(
    meta_use[, cols_meta, drop = FALSE],
    by = "cellID"
  ) %>%
  select(cellID, Krt5, Krt14, everything())

write.csv(
  expr_SCT_out,
  file = "./OutputTable/Palate_zhao_epi_Krt5_Krt14_SCT_expression.csv",
  row.names = FALSE,
  quote = FALSE
)

head(expr_SCT_out)
```


```{r}
cor.test(
  expr_RNA_out$Krt5,
  expr_RNA_out$Krt14,
  method = "spearman"
)

cor.test(
  expr_SCT_out$Krt5,
  expr_SCT_out$Krt14,
  method = "spearman"
)
```


