---
title: "Integrative Krt14 and Krt5 correlation plot"
author: "Huan Liu"
date: "2026-6-11"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  eval = FALSE,
  tidy.opts = list(width.cutoff = 120),
  message = TRUE,
  warning = FALSE,
  cache = TRUE,
  label.duplicate = "append"
)

```

This notebook records our approach to test the correlation between Krt14 and Krt5 in mouse embryonic palate epithelium, embryonic skin epithelium and young adult skin epithelium.

# 1 Load packages and datasets

```{r,load of packages and setup of the environment 1}
gc()
rm(list=ls())
library(readr)
library(dplyr)
library(purrr)
library(ggplot2)
library(patchwork)


```
# 2 Extract Krt14 and Krt5 from different scRNA object
## 2.1 Palate_zhao_epi

```{r}
Palate_zhao_epi <- readRDS(file = "./SeuratObject/Palate_zhao_epi_Celltype_marker_analysis.rds")

```
```{r}
genes_use <- c("Krt5", "Krt14")

## 检查 RNA 中是否存在
DefaultAssay(Palate_zhao_epi) <- "RNA"

genes_use[genes_use %in% rownames(Palate_zhao_epi[["RNA"]])]
genes_use[!genes_use %in% rownames(Palate_zhao_epi[["RNA"]])]
```

```{r}

Palate_zhao_epi <- NormalizeData(
  Palate_zhao_epi,
  assay = "RNA",
  verbose = FALSE
)

## 提取 RNA normalized expression
expr_RNA <- FetchData(
  Palate_zhao_epi,
  vars = genes_use,
  assay = "RNA",
  layer = "data"
)

expr_RNA$cellID <- rownames(expr_RNA)

## 加 metadata
meta_use <- Palate_zhao_epi@meta.data
meta_use$cellID <- rownames(meta_use)

cols_meta <- c("cellID", "Celltype", "Stage", "orig.ident")
cols_meta <- cols_meta[cols_meta %in% colnames(meta_use)]

expr_RNA_out <- expr_RNA %>%
  left_join(
    meta_use[, cols_meta, drop = FALSE],
    by = "cellID"
  ) %>%
  select(cellID, Krt5, Krt14, everything())

write.csv(
  expr_RNA_out,
  file = "./OutputTable/Palate_zhao_epi_Krt5_Krt14_RNA_expression.csv",
  row.names = FALSE,
  quote = FALSE
)

head(expr_RNA_out)
```


## 2.2 Young_skin
```{r}
skin_combined <- readRDS(file = "./young_integrated_celltype_20240125.rds")
```



### 2.2.1 Cluster information
```{r}
colnames(skin_combined@meta.data)

table(skin_combined$timepoint)
table(skin_combined$celltype)
```


```{r, initial plots}
#group_by_cluster
plot1 = DimPlot(skin_combined, reduction = "umap", pt.size=0.01, label = T, repel=T)
# group_by_sample
plot2 = DimPlot(skin_combined, reduction = "umap", pt.size=0.01, group.by = "timepoint")
plot3 = DimPlot(skin_combined, reduction = "umap", pt.size=0.01, label = T, split.by  = "timepoint")
# combine
plotc <- plot1 + plot2
ggsave("./skin_combined_umap.pdf", plot = plotc, width = 20, height = 10)
plot3
ggsave("./skin_combined_umap_split.pdf", plot = plot3, width = 24, height = 10)

DimPlot(skin_combined, reduction = "umap", pt.size=0.01, label = T)
DimPlot(skin_combined, reduction = "umap", pt.size=0.01, label = T,group.by = "orig.ident")

```

```{r,fig.height=6, fig.width=11}
FeaturePlot(skin_combined, features = c("Krt5"),
            cols = c("grey85",brewer.pal(9,"YlOrRd")),
             split.by = "timepoint",label = T,repel = T,
            ncol=2, pt.size = 0.3)
ggsave("./skin_combined_krt5_split.pdf", device = "pdf", width = 11, height = 5)

```

```{r,fig.height=6, fig.width=11}
FeaturePlot(skin_combined, features = c("Krt14"),
            cols = c("grey85",brewer.pal(9,"YlOrRd")),
             split.by = "timepoint",label = T,repel = T,
            ncol=2, pt.size = 0.3)
ggsave("./skin_combined_krt14_split.pdf", device = "pdf", width = 11, height = 5)

```


```{r}
p1 <- FeaturePlot(
  skin_combined,
  features = "Krt5",
  split.by = "timepoint",
  cols = c("grey85", "grey85"),
  pt.size = 0.3,
  label = TRUE,
  repel = TRUE,
  combine = TRUE
) + theme(legend.position = "right")

ggsave("./skin_combined_Krt5_split_1.pdf", p1, width=11, height=5)

```

```{r}
p_krt5 <- FeaturePlot(
  skin_combined,
  features = "Krt5",
  split.by = "timepoint",
  cols = c("grey85", brewer.pal(9, "YlOrRd")),
  pt.size = 0.3,
  label = TRUE,
  repel = TRUE,
  combine = TRUE        # 必须启用，否则不显示legend
) + 
  theme(legend.position = "right")

ggsave("./skin_combined_krt5_split_with_scalebar.pdf",
       p_krt5, width = 11, height = 5)

```


### 2.2.2 Subset Epithelium in uw
```{r}
skin_combined <- UpdateSeuratObject(skin_combined)

skin_epi_uw <- subset(
  skin_combined,
  subset = timepoint == "uw" & celltype == "Epithelium"
)
```

```{r}
colnames(skin_combined@meta.data)

table(skin_combined$timepoint)
table(skin_combined$celltype)
```
```{r}
skin_epi_uw <- subset(
  skin_combined,
  subset = timepoint == "uw" & celltype == "Epithelium"
)

skin_epi_uw

table(skin_epi_uw$timepoint)
table(skin_epi_uw$celltype)
```

```{r}


saveRDS(
  skin_epi_uw,
  file = "./SeuratObject/skin_combined_uw_Epithelium_subset.rds"
)
```

###2.2.3 Extract Krt14 and Krt5

Extract RNA
```{r}
dir.create("./OutputTable", showWarnings = FALSE, recursive = TRUE)

genes_use <- c("Krt5", "Krt14")

DefaultAssay(skin_epi_uw) <- "RNA"

## 检查基因是否存在
genes_use[genes_use %in% rownames(skin_epi_uw[["RNA"]])]
genes_use[!genes_use %in% rownames(skin_epi_uw[["RNA"]])]
```

```{r}
## 如果 RNA data layer 还没有 normalized data，先 NormalizeData
skin_epi_uw <- NormalizeData(
  skin_epi_uw,
  assay = "RNA",
  verbose = FALSE
)

expr_skin_RNA <- FetchData(
  skin_epi_uw,
  vars = genes_use,
  assay = "RNA",
  layer = "data"
)

expr_skin_RNA$cellID <- rownames(expr_skin_RNA)

meta_skin <- skin_epi_uw@meta.data
meta_skin$cellID <- rownames(meta_skin)

## 只保留你这个 object 里真实存在、后续有用的 metadata
cols_meta <- c(
  "cellID",
  "sample_id",
  "celltype",
  "timepoint",
  "orig.ident",
  "run",
  "version"
)

cols_meta <- cols_meta[cols_meta %in% colnames(meta_skin)]

expr_skin_RNA_out <- expr_skin_RNA %>%
  left_join(
    meta_skin[, cols_meta, drop = FALSE],
    by = "cellID"
  ) %>%
  mutate(Dataset = "mouse_skin_uw_Epithelium") %>%
  select(
    Dataset,
    cellID,
    sample_id,
    celltype,
    timepoint,
    Krt5,
    Krt14,
    everything()
  )

write.csv(
  expr_skin_RNA_out,
  file = "./OutputTable/mouse_skin_uw_Epithelium_Krt5_Krt14_RNA_expression.csv",
  row.names = FALSE,
  quote = FALSE
)

head(expr_skin_RNA_out)
```

```{r}
cor.test(
  expr_skin_RNA_out$Krt5,
  expr_skin_RNA_out$Krt14,
  method = "spearman"
)
```



# 3 Import RNA matrix

```{r}
## 1. Define input files
## =========================================================

file_info <- data.frame(
  scRNA_datasets = c(
    "Young_adult_mice_skin_epi",
    "Embryo_mice_skin_epi",
    "Embryo_mice_palate_epi"
  ),
  file_path = c(
    "./OutputTable/mouse_skin_uw_Epithelium_Krt5_Krt14_RNA_expression.csv",
    "./OutputTable/Adata-object_Epi_Krt5_Krt14_normalized_expression.csv",
    "./OutputTable/Palate_zhao_epi_Krt5_Krt14_RNA_expression.csv"
  ),
  stringsAsFactors = FALSE
)

file_info
```

```{r}
file_info$file_exists <- file.exists(file_info$file_path)

file_info
```

```{r}
## =========================================================
## 2. Read files
## =========================================================

expr_list <- list()

for (i in seq_len(nrow(file_info))) {
  
  dataset_i <- file_info$scRNA_datasets[i]
  file_i <- file_info$file_path[i]
  
  message("Reading: ", dataset_i)
  message("File: ", file_i)
  
  expr_list[[dataset_i]] <- read.csv(
    file_i,
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
  
  message("Dimension:")
  print(dim(expr_list[[dataset_i]]))
  
  message("Column names:")
  print(colnames(expr_list[[dataset_i]]))
  
  message("First rows:")
  print(head(expr_list[[dataset_i]]))
  
  message("--------------------------------------------------")
}
```

# 4 Merge three table
```{r}
## =========================================================
## Standardize three Krt5/Krt14 expression tables
## =========================================================

Young_adult_mice_skin_epi <- expr_list[["Young_adult_mice_skin_epi"]] %>%
  mutate(
    scRNA_datasets = "Young adult mice skin epithelium",
    cellID = cellID,
    group = timepoint,
    sample_id = sample_id
  ) %>%
  select(
    scRNA_datasets,
    cellID,
    Krt5,
    Krt14,
    group,
    sample_id,
    everything()
  )

Embryo_mice_skin_epi <- expr_list[["Embryo_mice_skin_epi"]] %>%
  mutate(
    scRNA_datasets = "Embryonic mice skin epithelium",
    cellID = cell_id,
    group = embryonic_age,
    sample_id = sample_id
  ) %>%
  select(
    scRNA_datasets,
    cellID,
    Krt5,
    Krt14,
    group,
    sample_id,
    everything()
  )

Embryo_mice_palate_epi <- expr_list[["Embryo_mice_palate_epi"]] %>%
  mutate(
    scRNA_datasets = "Embryonic mice palatal epithelium",
    cellID = cellID,
    group = Stage,
    sample_id = orig.ident
  ) %>%
  select(
    scRNA_datasets,
    cellID,
    Krt5,
    Krt14,
    group,
    sample_id,
    everything()
  )

Krt5_Krt14_expr_merged <- bind_rows(
  Young_adult_mice_skin_epi,
  Embryo_mice_skin_epi,
  Embryo_mice_palate_epi
)

Krt5_Krt14_expr_merged$scRNA_datasets <- factor(
  Krt5_Krt14_expr_merged$scRNA_datasets,
  levels = c(
    "Young adult mice skin epithelium",
    "Embryonic mice skin epithelium",
    "Embryonic mice palatal epithelium"
  )
)

head(Krt5_Krt14_expr_merged)
table(Krt5_Krt14_expr_merged$scRNA_datasets)
```
```{r}
write.csv(
  Krt5_Krt14_expr_merged,
  file = "./OutputTable/Krt5_Krt14_expression_merged_three_datasets.csv",
  row.names = FALSE,
  quote = FALSE
)
```

# 5 Calculate Spearman
```{r}
cor_summary_three <- Krt5_Krt14_expr_merged %>%
  group_by(scRNA_datasets) %>%
  summarise(
    n = n(),
    spearman_rho = cor(Krt5, Krt14, method = "spearman", use = "complete.obs"),
    pearson_r = cor(Krt5, Krt14, method = "pearson", use = "complete.obs"),
    .groups = "drop"
  )

cor_summary_three

write.csv(
  cor_summary_three,
  file = "./OutputTable/Krt5_Krt14_correlation_summary_three_datasets.csv",
  row.names = FALSE,
  quote = FALSE
)
```

```{r}
p_Krt5_Krt14_scatter_three <- ggplot(
  Krt5_Krt14_expr_merged,
  aes(x = Krt5, y = Krt14)
) +
  geom_point(
    aes(color = scRNA_datasets),
    size = 0.35,
    alpha = 0.45
  ) +
  facet_wrap(
    ~ scRNA_datasets,
    nrow = 1,
    scales = "free"
  ) +
  theme_classic() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 11),
    legend.position = "right"
  ) +
  labs(
    x = "Krt5 expression",
    y = "Krt14 expression",
    color = "scRNA-seq dataset"
  )

p_Krt5_Krt14_scatter_three
```

```{r}
dataset_colors <- c(
  "Young adult mice skin epithelium" = "#00BFC4",
  "Embryonic mice skin epithelium" = "#7CAE00",
  "Embryonic mice palatal epithelium" = "#F8766D"
)

p_Krt5_Krt14_scatter_three <- ggplot(
  Krt5_Krt14_expr_merged,
  aes(x = Krt5, y = Krt14)
) +
  geom_point(
    aes(color = scRNA_datasets),
    size = 0.35,
    alpha = 0.45
  ) +
  facet_wrap(
    ~ scRNA_datasets,
    nrow = 1,
    scales = "free"
  ) +
  scale_color_manual(values = dataset_colors) +
  theme_classic() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 11),
    legend.position = "right"
  ) +
  labs(
    x = "Krt5 expression",
    y = "Krt14 expression",
    color = "scRNA-seq dataset"
  )

p_Krt5_Krt14_scatter_three
```


```{r}
ggsave(
  "./OutputFigure/Krt5_Krt14_scatter_three_scRNA_datasets.pdf",
  plot = p_Krt5_Krt14_scatter_three,
  width = 9,
  height = 3.2,
  units = "in"
)

```

```{r}
library(ggplot2)
library(dplyr)

## correlation label
cor_label_df <- cor_summary_three %>%
  mutate(
    label = paste0(
      "Spearman \u03c1 = ",
      round(spearman_rho, 2),
      "\nPearson r = ",
      round(pearson_r, 2)
    )
  )

p_Krt5_Krt14_scatter_three_lm <- ggplot(
  Krt5_Krt14_expr_merged,
  aes(x = Krt5, y = Krt14)
) +
  geom_point(
    aes(color = scRNA_datasets),
    size = 0.35,
    alpha = 0.35
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "black",
    linewidth = 0.6
  ) +
  geom_text(
    data = cor_label_df,
    aes(
      x = Inf,
      y = Inf,
      label = label
    ),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.2,
    size = 3
  ) +
  facet_wrap(
    ~ scRNA_datasets,
    nrow = 1,
    scales = "free"
  ) +
  scale_color_manual(
    values = c(
      "Young adult mice skin epithelium" = "#00BFC4",
      "Embryonic mice skin epithelium" = "#7CAE00",
      "Embryonic mice palatal epithelium" = "#F8766D"
    )
  ) +
  theme_classic() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 11),
    legend.position = "none"
  ) +
  labs(
    x = "Krt5 expression",
    y = "Krt14 expression"
  )

p_Krt5_Krt14_scatter_three_lm
```

```{r}
ggsave(
  "./OutputFigure/Krt5_Krt14_scatter_three_scRNA_datasets_lm_spearman.pdf",
  plot = p_Krt5_Krt14_scatter_three_lm,
  width = 9,
  height = 3.2,
  units = "in"
)
```



```{r}
p_Krt5_Krt14_scatter_three_loess <- ggplot(
  Krt5_Krt14_expr_merged,
  aes(x = Krt5, y = Krt14)
) +
  geom_point(
    aes(color = scRNA_datasets),
    size = 0.35,
    alpha = 0.35
  ) +
  geom_smooth(
    method = "loess",
    se = FALSE,
    color = "black",
    linewidth = 0.6
  ) +
  geom_text(
    data = cor_label_df,
    aes(
      x = Inf,
      y = Inf,
      label = label
    ),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.2,
    size = 3
  ) +
  facet_wrap(
    ~ scRNA_datasets,
    nrow = 1,
    scales = "free"
  ) +
  scale_color_manual(
    values = c(
      "Young adult mice skin epithelium" = "#00BFC4",
      "Embryonic mice skin epithelium" = "#7CAE00",
      "Embryonic mice palatal epithelium" = "#F8766D"
    )
  ) +
  theme_classic() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 11),
    legend.position = "none"
  ) +
  labs(
    x = "Krt5 expression",
    y = "Krt14 expression"
  )

p_Krt5_Krt14_scatter_three_loess
```

```{r}
ggsave(
  "./OutputFigure/Krt5_Krt14_scatter_three_scRNA_datasets_loess_spearman.pdf",
  plot = p_Krt5_Krt14_scatter_three_loess,
  width = 9,
  height = 3.2,
  units = "in"
)
```


# 6 Correlation plot (Supplementary Fig 1f)
```{r}
cor_summary_three <- Krt5_Krt14_expr_merged %>%
  group_by(scRNA_datasets) %>%
  summarise(
    n = n(),
    spearman_rho = cor(Krt5, Krt14, method = "spearman", use = "complete.obs"),
    pearson_r = cor(Krt5, Krt14, method = "pearson", use = "complete.obs"),
    .groups = "drop"
  )

cor_summary_three
```

```{r}
p_Krt5_Krt14_scatter_three <- ggplot(
  Krt5_Krt14_expr_merged,
  aes(x = Krt5, y = Krt14)
) +
  geom_point(
    aes(color = scRNA_datasets),
    size = 0.35,
    alpha = 0.40
  ) +
  geom_text(
    data = cor_label_df,
    aes(
      x = Inf,
      y = Inf,
      label = label
    ),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.2,
    size = 3.5
  ) +
  facet_wrap(
    ~ scRNA_datasets,
    nrow = 1,
    scales = "free"
  ) +
  scale_color_manual(
    values = c(
      "Young adult mice skin epithelium" = "#00BFC4",
      "Embryonic mice skin epithelium" = "#7CAE00",
      "Embryonic mice palatal epithelium" = "#F8766D"
    )
  ) +
  theme_classic() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 11),
    legend.position = "none"
  ) +
  labs(
    x = "Krt5 expression",
    y = "Krt14 expression"
  )

p_Krt5_Krt14_scatter_three
```

```{r}
ggsave(
  "./OutputFigure/Krt5_Krt14_scatter_three_scRNA_datasets_spearman.pdf",
  plot = p_Krt5_Krt14_scatter_three,
  width = 9,
  height = 3.2,
  units = "in"
)
```